---
title: "Getting Started with MethylTraits"
author: "Andy Madrid"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with MethylTraits}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
data_path <- "/media/data/dormant/Epigenetic_Age/bs.pasc3.rda"
data_available <- file.exists(data_path)
knitr::opts_chunk$set(eval = data_available)
```

```{r, echo = FALSE, results = "asis"}
if (!data_available) {
  cat(
    "**Note:** this vignette was built without access to the internal validation ",
    "dataset (`bs.pasc3.rda`) used to demonstrate every function below, so the ",
    "code chunks were not evaluated. Every example runs the same way on any ",
    "`BSseq` object -- just substitute your own."
  )
}
```

## Introduction

MethylTraits works directly on `BSseq` objects (from the
[bsseq](https://bioconductor.org/packages/bsseq/) package) and covers two
complementary jobs:

1. **Discovery** -- find and annotate regulatory methylation features across a
   cohort: imprinting control regions (`findICRs`, `annotateICRs`) and stochastic
   epigenetic mutations (`findSEMs`, `annotateSEMs`), plus gene-level burden
   association testing (`testSEMBurden`, `testSEMGenes`) and GO/KEGG enrichment
   (`goSEMs`) on top of them.
2. **Prediction** -- predict sample-level traits directly from WGBS methylation:
   genetic sex and sex-chromosome karyotype (`predictSex`), self-reported
   race/ethnicity (`predictRace`), PASC/long-COVID status (`predictPASC`),
   epigenetic age via a two-tier ensemble (`predictAge`), and 450K/EPIC array
   emulation for cross-platform work (`emulateArray`).

A third function, `liftBuild`, doesn't belong to either group: it's a
preprocessing utility that converts a `BSseq` object's coordinates between
hg19 and hg38, for when your data isn't already on the build a given function
expects.

Nothing in MethylTraits requires you to train anything yourself: `predictSex` is
purely coverage/copy-number based (no training data needed at all), and
`predictRace`/`predictPASC`/`predictAge`/`emulateArray` ship with pre-trained
models bundled inside the package (`system.file("models", ..., package =
"MethylTraits")`) -- you only ever supply a `BSseq` object.

## Installation

```r
# Bioconductor dependencies used throughout the package
if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
BiocManager::install(c("bsseq", "GenomicRanges", "IRanges", "S4Vectors", "GenomeInfoDb"))

# from a locally built source tarball
install.packages("MethylTraits_0.1.0.tar.gz", repos = NULL, type = "source")
```

Most functions also need one or more packages listed under `Suggests` in
`DESCRIPTION` (e.g. `glmnet` for `predictRace`/`predictPASC`, `xgboost`/`ranger`
for `predictAge`, `minfi` for `emulateArray`) -- install these as needed; each
function will tell you at call time if something required is missing.

## Example data

The examples below use a small subset of samples from an internal WGBS
validation cohort (`bs.pasc3.rda`, 68 samples x ~28.9M CpGs genome-wide) that was
used throughout this package's development to validate every function against
real data. Substitute your own `BSseq` object anywhere `bs` appears below.

```{r load-data}
suppressPackageStartupMessages(library(bsseq))
library(MethylTraits)

load(data_path)
bs <- bs.pasc3
rm(bs.pasc3)

# a 3-sample subset keeps the *prediction* examples fast; predictSex/predictRace/
# predictPASC/predictAge/emulateArray all work identically on cohorts of any size
bs_small <- bs[, 1:3]
sampleNames(bs_small) <- pData(bs_small)$"Sample ID"
bs_small
```

## Genome build conversion

`bs.pasc3.rda` is already on hg38, the build every other function in this
package expects by default. If your own data is aligned to hg19 instead (or
you need to move hg38 data down to hg19, e.g. to match an older annotation
resource), `liftBuild` converts a `BSseq` object's coordinates between the two
via a UCSC chain file (fetched once and cached locally through
`AnnotationHub`). CpGs with no equivalent locus in the target build are
dropped, and `genome(bs)` is updated to reflect the new build.

```{r lift-build}
bs_small_hg38 <- bs_small[as.character(seqnames(bs_small)) %in% c("chr21", "chr22"), ]
bs_hg19 <- liftBuild(bs_small_hg38, current = "hg38", new = "hg19")
dim(bs_small_hg38)
dim(bs_hg19)
unique(genome(bs_hg19))
```

## Part 1: Discovery

The discovery functions look for signal across CpGs genome-wide, so their
runtime scales with how much of the genome you give them. To keep this vignette
fast we restrict to chr21+chr22 (~990K of the ~28.9M CpGs) and use the full
68-sample cohort, since `findSEMs`'s outlier logic needs a reasonably large
cohort to be meaningful; the same calls work genome-wide on your own data by
simply passing the full `bs` object.

```{r discovery-subset}
bs_demo <- bs[as.character(seqnames(bs)) %in% c("chr21", "chr22"), ]
sampleNames(bs_demo) <- pData(bs_demo)$"Sample ID" # guarantee unique sample IDs
bs_demo
```

### Imprinting control regions (ICRs)

`findICRs` scans for windows of consecutive, densely-spaced CpGs that are
consistently hemimethylated (or monoallelic, in gamete tissue) across the
cohort -- the signature of genomic imprinting -- following the approach of
Jima et al. 2022.

```{r find-icrs}
icrs <- findICRs(bs_demo, minCov = 4, minSample = 0.5, ncores = 4)
head(icrs)
```

`annotateICRs` then assigns each candidate window to its nearest gene/feature
and flags whether it overlaps a previously published ICR (Jima et al. 2022):

```{r annotate-icrs}
icrs_annotated <- annotateICRs(icrs, genomeBuild = "hg38")
head(icrs_annotated)
```

### Stochastic epigenetic mutations (SEMs)

`findSEMs` flags, per sample and per CpG, methylation values that are extreme
outliers (Tukey-fence based) relative to the rest of the cohort at that CpG --
these are stochastic epigenetic mutations. `saveIndSEMs = TRUE` additionally
writes out the long-format per-event table (`sampleSEMs.rda`) that
`annotateSEMs` needs.

```{r find-sems}
sample_sems <- findSEMs(
  bs_demo, minCov = 10, minSamples = 50,
  saveIndSEMs = TRUE, saveDir = tempdir(), ncores = 4
)
head(sample_sems)
```

`annotateSEMs` maps every SEM-carrying CpG to its nearest gene and aggregates
into a genes x samples SEM-count matrix, ready for association testing:

```{r annotate-sems}
load(file.path(tempdir(), "sampleSEMs.rda")) # loads cpgSEMs
sem_annotated <- annotateSEMs(
  cpgSEMs = cpgSEMs, sampleNames = sampleNames(bs_demo), genomeBuild = "hg38",
  universeCpGs = bs_demo
)
str(sem_annotated, max.level = 1)
head(sem_annotated$cpgGeneMap)
```

`testSEMBurden` asks whether total per-sample SEM burden differs across a
grouping variable (e.g. diagnosis, PASC status, sex) -- a simple omnibus check
analogous to a "Global-EML" metric:

```{r test-sem-burden}
group <- pData(bs_demo)$Sex
burden <- testSEMBurden(sample_sems, group = group)
burden$test
burden$summary
```

`testSEMGenes` tests each individual gene's SEM count for association with the
same grouping variable, modeled with `edgeR` (handles the sparse, over-dispersed
count data properly and supports covariates/>2 groups):

```{r test-sem-genes}
gene_test <- testSEMGenes(sem_annotated$geneMat, group = group)
head(gene_test$results)
```

Finally, `goSEMs` runs GO/KEGG over-representation on the genes that came out
significant, against a proper background gene universe rather than the whole
genome:

```{r go-sems}
sig_genes <- gene_test$results$gene[gene_test$results$FDR < 0.25]
if (length(sig_genes) > 0) {
  go_res <- goSEMs(sig_genes, universe = sem_annotated$universeGenes, runKEGG = FALSE)
  head(as.data.frame(go_res$GO))
} else {
  message("No genes passed the demo FDR cutoff on this small chr21+chr22 subset -- ",
          "expected on a 2-chromosome slice; try the full genome on real data.")
}
```

## Part 2: Prediction

These functions predict one value per sample and run comfortably on a handful
of samples, so we go back to the fast 3-sample `bs_small` object and use the
full genome (needed for `predictSex`'s coverage signal, and for reliable
coordinate/CpG matching against each predictor's bundled panel).

### Genetic sex and sex-chromosome karyotype

`predictSex` needs no training data: it estimates each sample's copy number on
chrX/chrY relative to the autosomal baseline from raw coverage alone, and
classifies against theoretical karyotype centroids (XX, XY, X0/Turner,
XXY/Klinefelter, XXX/triple X, XYY).

```{r predict-sex}
sex_res <- predictSex(bs_small)
sex_res[, c("sample", "karyotype", "predictedSex")]
```

Passing `knownSex` for any samples with independently known sex both
recalibrates the reference centroids to this assay's own depth ratios and
reports predicted-vs-known concordance -- useful as a standalone sample-swap QC
check even if you don't care about sex prediction per se:

```{r predict-sex-calibrated}
sex_res_cal <- predictSex(bs_small, knownSex = pData(bs_small)$Sex)
attr(sex_res_cal, "calibrated")
```

### Race/ethnicity

```{r predict-race}
race_res <- predictRace(bs_small)
race_res[, c("sample", "predictedRace", "maxProb")]
```

### PASC (long-COVID) status

```{r predict-pasc}
pasc_res <- predictPASC(bs_small)
pasc_res[, c("sample", "prob_enet", "predictedStatus")]
```

`predictPASC` prints a calibration caveat at runtime: its 0.5 threshold was
only validated on the training platform (WGBS-emulated EPIC data), so on real
input `prob_enet`'s relative ranking across your own samples is more
trustworthy than the absolute `predictedStatus` cutoff.

### Epigenetic age

```{r predict-age}
age_res <- predictAge(bs_small)
age_res[, c("sample", "wgbs_only_stack", "hybrid_stack")]
pData(bs_small)$Age # known chronological ages, for comparison
```

`wgbs_only_stack` (Tier 1) is the primary estimate, built purely from WGBS
data; `hybrid_stack` (Tier 2) additionally incorporates published clocks and is
provided as a secondary, applied estimate.

### Array emulation

`emulateArray` converts WGBS CpGs to their corresponding probe-level beta
values on a 450K/EPICv1/EPICv2 array layout, for cross-platform comparisons or
to run array-native tools (methylCIPHER clocks, DunedinPACE) that expect array
input.

```{r emulate-array, warning = FALSE}
# warning = FALSE: methylCIPHER's own clock formulas emit benign vector-
# recycling warnings on small inputs; harmless, just noisy in a vignette
arr <- emulateArray(bs_small, array = "EPICv1", runClocks = TRUE)
dim(arr$beta)
arr$ageEstimates[, 1:6]
```

## Session info

```{r session-info}
sessionInfo()
```
